---
title: "Seurat_tutorial"
author: "Meryem Stroosma"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    latex_engine: xelatex

# Use colours for hyperlinks
colorlinks: true
linkcolor: purple
urlcolor: purple
citecolor: purple

header-includes:
  - \usepackage{fontspec}
  - \usepackage{fancyvrb}
  - \usepackage{placeins}
  - \usepackage{float} 
  - \setmainfont{DejaVu Sans Condensed}
  - \setsansfont{DejaVu Sans}
  - \setmonofont{DejaVu Sans Mono}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{fontsize=\footnotesize,commandchars=\\\{\}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Content
- [Introduction](#introduction)
- [Purpose](#purpose)
- [Data](#data)
- [Packages](#packages)
- [Loading the data](#loading-the-dataset-and-creating-a-seurat-object)


## Introduction
This RMarkdown contains a brief summary of the R package Seurat. Seurat is used for the analysis of single-cell RNA sequencing data, providing tools for data preprocessing, quality control, normalization, dimensionality reduction, clustering, differential expression, and cell-type annotation. Besides analyzing a dataset Seurat can also be used to create a variaty of plots to visualize the data. 
The summary in this RMarkdown is based on the [Seurat guided clustering tutorial](https://satijalab.org/seurat/articles/pbmc3k_tutorial).

## Purpose
The purpose of this RMarkdown is to understand the different functionalities of Seurat, and to learn how to use Seurat 
for analyzing the [GSE138852](https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE138852&format=file&file=GSE138852%5Fcounts%2Ecsv%2Egz) count matrix.

## Data
The data that is used in this RMarkdown can be accessed [here](https://cf.10xgenomics.com/samples/cell/pbmc3k/pbmc3k_filtered_gene_bc_matrices.tar.gz). This is a dataset of Peripheral Blood Mononuclear Cells (PBMC) freely available from 10X Genomics. This dataset is also used in the [Seurat guided clustering tutorial](https://satijalab.org/seurat/articles/pbmc3k_tutorial).

### Packages
Prior to analysis several packages need to be loaded. The required packages are the following:

 * ```dplyr```: this package is used by Seurat for data manipulation.
 * ```Seurat```: this package needs to be loaded prior to running any Seurat specific functions.
 * ```patchwork```: the patchwork package can be used to combine several plots into a single figure.
 * ```here```: this package is used to make sure the file paths are reproducible.

```{r Loading packages, echo=FALSE, include=FALSE}
# Load the required packages
library(dplyr)
library(Seurat)
library(patchwork)
library(here)
library(ggplot2)
```


### Loading the dataset and creating a Seurat object
In order to perform analysis with Seurat, the data first needs to be loaded. This is done with ```Read10X()``` which is a Seurat function that can be used to load 10X genomics files such as matrices and .tsv files.
After loading the dataset, the function ```CreateSeuratObject()``` is used to create a Seurat object that contains the raw data. In this object several filters can be applied or altered, which is shown in the following sections.
```{r Load the dataset, include=FALSE}
# Load the PBMC dataset
pbmc.data <- Read10X(data.dir = here("seurat", "filtered_gene_bc_matrices", "hg19"))

# Initialize the Seurat object with the raw (non-normalized data).
pbmc <- CreateSeuratObject(counts = pbmc.data, project = "pbmc3k", min.cells = 3, min.features = 200)
pbmc

```

### Pre-processing of the dataset
To make sure the data that will be analyzed is reliable it is important that low quality cells are removed from the dataset. In this case the quality doesn't refer to a Phred score (read quality), but to the amount of unique features per cell, the percentage of mitochondrial genes and the total number of molecules per cell;

 - The number of unique features per cell are relevant since too few features compared to the rest of the cells could mean that sequencing of this particular cell was insufficient or that the droplet did not contain a cell. If the number of features is higher it could mean that two or more cells were captured in the same droplet. In most cases a feature is a gene, but this doesn't always have to be the case, sometimes this might be regulatory RNA instead. 
 - The number of counts says something about the number of different molecules or Unique Molecular Identifiers (UMIs) in one cell. If this number is too low it could mean that this is a poor capture and if this number is too high it could mean that there might be more then one cell in a droplet.
 - The percentage of mitochondrial genes are a good indication about the general health of the cell, if this percentage is high it usually means that the cell is dying or stressed. Since it might not be representative for the condition that is analysed it is better to filter this out. 
 
The first step in pre-processing is adding a column to the dataframe with the percentage of mitochondrial genes. To give a general idea of what the dataframe looks like after adding this column the first five rows are shown in the chunk below.
In the first column the cell barcodes are shown, the second column contains the original identity of the sample the third, fourth and fifth column contain the number of the number of counts, unique genes and the percentage of mitochondrial genes. 
 
```{r Pre-processing of the dataset, echo=FALSE}
# Add a column with [[ to the Seurat object containing info about the percentage of mitochondrial genes, PercentageFeatureSet searches for genes that start with MT in the pbmc dataset.
pbmc[["percent_mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")

# Show the first 5 lines of the meta data frame of Seurat object pbmc
head(pbmc@meta.data, 5)
```
 
The dataframe can also be visualised in a plot, in figure 1 a violin plot is shown of the following metrics;

 - __nFeature_RNA__: this plot shows the unique number of genes
 - __nCount_RNA__: this plot shows the total amount of RNA molecules detected
 - __percent_mt__: this plot shows the percentage of mitochondrial genes
In the violin plots in figure 1 each dot represents the datapoint for one single cell, on the x-axis the identity is shown, on the y-axis the amount is shown of the metric that is plotted.

\FloatBarrier
```{r Visualize dataframe, echo=FALSE, warning=FALSE, fig.cap="Violin plot of QC-metrics, in the first plot the unique number of genes per cell are visualised, in the second plot the amount of RNA molecules is shown and in the last plot the percentage of mitochondrial genes is shown"}
# Visualize dataframe as a violin plot
VlnPlot(pbmc, features = c("nFeature_RNA", "nCount_RNA", "percent_mt"), ncol = 3)
```
\FloatBarrier
With the function ```FeatureScatter()``` a scatterplot is generated of two QC metrics. The first argument should contain the Seurat object where the QC metrics are stored. Then the x-axis should be defined with the argument ```feature1 =```, in figure 2 this was defined as nCount_RNA. The y-axis can be defined in a similar way using ```feature2 =```, in the left plot in figure 2 this was defined as *percent_mt* and in the right plot this was defined as *nFeature_RNA*. The title of the plot is the correlation value, this could be changed by the argument plot.cor, which can be set to TRUE or FALSE. ```FeatureScatter()``` also has several arguments that could be used to modify the plots, for example if the dataset contains cells with different identities these could be grouped in different colours with argument group.by. 

\FloatBarrier
```{r Visualize dataframe as scatterplot, echo=FALSE, warning =FALSE, fig.cap="Scatterplot to visualise a relationship between the percentage of mitochondrial RNA versus the total amount of RNA and to visualize the relationship between the total amount of features and the total amount of RNA"}

# Visualize dataframe in scatterplots
plot1 <- FeatureScatter(pbmc,
                        feature1 = "nCount_RNA",
                        feature2 = "percent_mt",
                        plot.cor = FALSE) +
  ggtitle("Mitochondrial percentage\nverus RNA counts") +
  xlab("Total RNA counts") +
  ylab("Percentage MT genes") +
  theme_classic()
plot2 <- FeatureScatter(pbmc, 
                        feature1 = "nCount_RNA",
                        feature2 = "nFeature_RNA",
                        plot.cor = FALSE) +
  ggtitle("RNA features\nversus RNA counts") +
  xlab("Total RNA counts") +
  ylab("RNA features") +
  theme_classic()
plot1 + plot2
```
\FloatBarrier

The plots in figure 1 and figure 2 contain the unfiltered data of the pbmc dataframe. In order to 











