---
title: "Seurat_tutorial"
author: "Meryem Stroosma"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    latex_engine: xelatex

# Tell Pandoc/LaTeX to use colored links
colorlinks: true
linkcolor: purple
urlcolor: purple
citecolor: purple

header-includes:
  - \usepackage{fontspec}
  - \usepackage{fancyvrb}
  - \setmainfont{DejaVu Sans Condensed}
  - \setsansfont{DejaVu Sans}
  - \setmonofont{DejaVu Sans Mono}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{fontsize=\footnotesize,commandchars=\\\{\}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
This RMarkdown contains a brief summary of the R package Seurat. Seurat is used for the analysis of single-cell RNA sequencing data, providing tools for data preprocessing, quality control, normalization, dimensionality reduction, clustering, differential expression, and cell-type annotation. Besides analyzing a dataset Seurat can also be used to create a variaty of plots to visualize the data. 
The summary in this RMarkdown is based on the [Seurat guided clustering tutorial](https://satijalab.org/seurat/articles/pbmc3k_tutorial).

## Purpose
The purpose of this RMarkdown is to understand the different functionalities of Seurat, and to learn how to use Seurat 
for analyzing the [GSE138852](https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE138852&format=file&file=GSE138852%5Fcounts%2Ecsv%2Egz) count matrix.

## Data
The data that is used in this RMarkdown can be accessed [here](https://cf.10xgenomics.com/samples/cell/pbmc3k/pbmc3k_filtered_gene_bc_matrices.tar.gz). This is a dataset of Peripheral Blood Mononuclear Cells (PBMC) freely available from 10X Genomics. This dataset is also used in the [Seurat guided clustering tutorial](https://satijalab.org/seurat/articles/pbmc3k_tutorial).

### Packages
Prior to analysis several packages need to be loaded. The required packages are the following:

* __dplyr__: this package is used by Seurat for data manipulation.  
* __Seurat__: this package needs to be loaded prior to running any Seurat specific functions.
* __patchwork__: the patchwork package can be used to combine several plots into a single figure.
* __here__: this package is used to make sure the file paths are reproducible.

```{r Loading packages, echo=FALSE, include=FALSE}
# Load the required packages
library(dplyr)
library(Seurat)
library(patchwork)
library(here)
```


### Loading the dataset and creating a Seurat object
```{r Load the dataset, include=FALSE}
# Load the PBMC dataset
pbmc.data <- Read10X(data.dir = here("seurat", "filtered_gene_bc_matrices", "hg19"))

# Initialize the Seurat object with the raw (non-normalized data).
pbmc <- CreateSeuratObject(counts = pbmc.data, project = "pbmc3k", min.cells = 3, min.features = 200)
pbmc

```
