---
title: "02 Identification of different celltypes in Seurat"
author: "Meryem Stroosma"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    latex_engine: xelatex
    fig_caption: true

# Use colours for hyperlinks
colorlinks: true
linkcolor: purple
urlcolor: purple
citecolor: purple

header-includes:
  - \usepackage{fontspec}
  - \usepackage{fancyvrb}
  - \usepackage[section]{placeins}
  - \usepackage{float} 
  - \usepackage{blindtext}
  - \setmainfont{DejaVu Sans Condensed}
  - \setsansfont{DejaVu Sans}
  - \setmonofont{DejaVu Sans Mono}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{fontsize=\footnotesize,commandchars=\\\{\}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Contents
 - [Introduction](#introduction)
 - [Research question](#research-question)
 - [Packages and requirements](#packages-and-requirements)
 - [About the count matrix](#about-the-count-matrix)
 - [Loading the dataset and creating a Seurat object](#loading-the-dataset-and-creating-a-seurat-object)
 - [Pre-processing of the dataset](#pre-processing-of-the-dataset)
 - [Normalization of the dataset](#normalization-of-the-dataset)
 - [Conclusion](Conclusion)
 - [Discussion](Discussion)

## Introduction
This RMarkdown describes an analysis of the single-nucleus RNA-sequencing count matrix from [GSE138852](https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE138852&format=file&file=GSE138852%5Fcounts%2Ecsv%2Egz) using Seurat. The dataset consists of a count matrix containing raw UMI counts derived from snRNAseq of human entorhinal cortex nuclei. Quality control of the nuclei was already performed according to the information provided in the article by Grubman _et al_., with low-quality cells removed during prior filtering steps.

The scope of this RMarkdown includes normalization of the count matrix, identification of highly variable features, and dimensionality reduction. Cell-type annotation is performed using BRETIGEA but is not covered within this RMarkdown.

## Research question
For this project the main research question is _"How can a reliable and reproducible analysis pipeline be set up to identify cell type-specific differential gene expression in Alzheimer's disease using a filtered expression matrix generated from snRNA-seq data?"_. Initially the research question was aimed at creating the entire pipeline with raw sequencing data as a starting point however, the publicly available data was lacking information on cellbarcodes and UMIs. Therefore the expression matrix published for GSE138852 was used as a starting point.

The sub-question(s) that is answered in this RMarkdown is as followed;

 - Can the most variable features be identified with the filtered expression count matrix as input? 
 - How should the count matrix be used, should the groups be separated (healthy vs. Alzheimers disease)? 

## Packages and Requirements
In order to run the analysis in Seurat a few packages need to be installed first. Packages and package versions are managed using ```renv```, instructions for setting up the R project with ```renv``` are given in the [README](https://github.com/ProjecticumDlerpDs/ad_transcriptomics_dge/blob/main/README.md) of the GitHub repository *ad_transcriptomics_dge*. 

```{r Load_packages, echo=FALSE, include=FALSE}
# Load the required packages
library(dplyr)
library(Seurat)
library(patchwork)
library(here)
library(ggplot2)
```

## About the count matrix
The count matrix was downloaded from the GEO-database. The count matrix is a table which stores the gene expression measurements of all measured nuclei. Usually the matrix contains the raw expression measurements but in the case of GSE138852 some filters were already applied, the matrix contains;

 - 13214 nuclei (before filtering 14876 nuclei were measured)
 - Only nuclei with less then 10% of the total features being mitochondrial
 - None of the features that are related to postmortem interval
 - Only cells that are within the 5th - 95th percentile regarding gene and UMI count. 
 
## Loading the dataset and creating a Seurat object
The count matrix was stored in a comma separated values file (.csv-file). To be able to use this file for creating a Seurat object ```read.csv()``` was used to create a data.frame first. Then the data.frame was used as input for the function ```CreateSeuratObject()```. 

```{r Load_data_set, include=FALSE}
# Load the GSE138852 countmatrix as a data.frame
count_matrix <- read.csv(file = here("raw_data", "GSE138852_counts.csv"),
  row.names = 1,
  check.names = FALSE
  )

# Initialize the Seurat object with the pre-filtered count matrix
count_matrix <- CreateSeuratObject(counts = count_matrix, project = "GSE138852")
```

## Pre-processing of the dataset
Since the count matrix had already been pre-processed as described in "[About the count matrix](#About-the-count-matrix)", no additional pre-processing is performed at this point. However to check if the dataset was indeed pre-processed as stated in the article by Grubman _et al_. a column was added containing the percentage of mitochondrial features. A violin plot was generated and visualized in figure 1. 

```{r Preprocessing_of_data_set, echo=FALSE}
# Add a column containing the percentage of genes starting with MT (mitochondrial).
count_matrix[["Percent_mt"]] <- PercentageFeatureSet(count_matrix, pattern = "^MT-")
```

```{r Visualize_dataframe, echo=FALSE, warning=FALSE, fig.cap="Violin plots of quality-control metrics. Shown are the number of detected genes per nucleus (nFeature_RNA), total UMI counts per nucleus (nCount_RNA), and the percentage of mitochondrial gene expression (Percent_mt). The distributions of these plots are as expected with prior quality-control filtering as described by Grubman et al.", fig.pos='!htbp'}
# Visualize dataframe as a violin plot
VlnPlot(count_matrix, features = c("nFeature_RNA", "nCount_RNA", "Percent_mt"), ncol = 3)
```

## Normalization of the dataset

```{r Normalize_data, include=FALSE, warning=FALSE}
# To normalize the data count_matrix is replaced by a new object for count_matrix in which the data is normalized by using NormalizeData with the default settings LogNormalize and scale.factor=10000.
count_matrix <- NormalizeData(count_matrix) 
```

After normalizing the dataset ```FindVariableFeatures()``` was used to detect the most variable features. The selection method that was used is ```vst```, and ```nFeatures``` was set at 2000. A scatter plot of the output of ```FindVariableFeatures()``` generated and visualized in figure 2. 

```{r Find_variable_features, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Scatterplot of variable features.", fig.pos='!htbp', fig.height=7}
# Find variable features 
count_matrix <- FindVariableFeatures(count_matrix, selection.method = "vst", nfeatures = 2000)
# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(count_matrix), 10)
# plot variable features without and with labels (with LabelPoints())
plot1 <- VariableFeaturePlot(count_matrix)+
  ggtitle("Features of GSE138852 dataset") +
  xlab("Average expression") +
  ylab("Standardized variance") +
  theme_classic()+
    theme(
    text = element_text(size = 10),        
    plot.title = element_text(size = 11),  
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 9),
    legend.position = "bottom")
plot1
    
```

## Conclusion
### Misschien dat dit result moet worden 
In order to check if the count matrix contained the data as described by the article by Grubman _et al_. and the GEO-database (Accession number: GSE138852) the data was plotted in three violin plots (figure 1). These violin plots confirm that the filtering was applied as stated in the article by Grubman et al., most nuclei show between 400 to 800 detected features, the amount of total RNA molecules detected is between 500 - 1000 for most nuclei and the percentage of mitochondrial features is around 1,5% with no values >10%. 

