---
title: "02 Identification of different celltypes in Seurat"
author: "Meryem Stroosma"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    latex_engine: xelatex
    fig_caption: true

# Use colours for hyperlinks
colorlinks: true
linkcolor: purple
urlcolor: purple
citecolor: purple

header-includes:
  - \usepackage{fontspec}
  - \usepackage{fancyvrb}
  - \usepackage[section]{placeins}
  - \usepackage{float} 
  - \usepackage{blindtext}
  - \setmainfont{DejaVu Sans Condensed}
  - \setsansfont{DejaVu Sans}
  - \setmonofont{DejaVu Sans Mono}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{fontsize=\footnotesize,commandchars=\\\{\}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Load_packages, echo=FALSE, include=FALSE}
# Load the required packages
library(dplyr)
library(Seurat)
library(patchwork)
library(here)
library(ggplot2)
```

## Introduction
In this RMarkdown the FastQC and multiQC analysis of the .fastq-files from [GSE138852](https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE138852&format=file&file=GSE138852%5Fcounts%2Ecsv%2Egz) is described. These .fastq-files contain the raw sequencing data and the phred-scores generated on the Illumina NextSeq 500 high-output lane with the 10x Genomics Chromium Single Cell 3′ Library & Gel Bead Kit v2. This markdown explains how identification of different cell types is be done and what the conclusion and discussion is in regards to the research question.

## Research question
For this project the main research question is _"How can a reliable and reproducible analysis pipeline be established to identify cell-type-specific differential gene expression from snRNA-seq data in Alzheimer’s disease?"_, the second step XXXXX

## Packages and Requirements
In order to run the scripts for analysis in Seurat a few packages need to be installed first, for analysis an environment was created in miniconda with the command line prompt;
XXXXXXXX


### Loading the dataset and creating a Seurat object
In order to perform analysis with Seurat, the data first needs to be loaded. This is done with ```Read10X()``` which is a Seurat function that can be used to load 10X genomics files such as matrices and .tsv files.
After loading the dataset, the function ```CreateSeuratObject()``` is used to create a Seurat object that contains the raw data. In this object several filters can be applied or altered, which is shown in the following sections.
```{r Load_data_set, include=FALSE}
# Load the GSE138852 countmatrix
count_matrix <- read.csv(file = here("raw_data", "GSE138852_counts.csv"),
  row.names = 1,
  check.names = FALSE
  )

# Initialize the Seurat object with the pre-filtered count matrix
count_matrix <- CreateSeuratObject(counts = count_matrix, project = "GSE138852")
```

