---
title: "Identification most variable features in Seurat"
author: "Meryem Stroosma"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    latex_engine: xelatex
    fig_caption: true

# Use colours for hyperlinks
colorlinks: true
linkcolor: purple
urlcolor: purple
citecolor: purple

header-includes:
  - \usepackage{fontspec}
  - \usepackage{fancyvrb}
  - \usepackage[section]{placeins}
  - \usepackage{float} 
  - \usepackage{blindtext}
  - \setmainfont{DejaVu Sans Condensed}
  - \setsansfont{DejaVu Sans}
  - \setmonofont{DejaVu Sans Mono}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{fontsize=\footnotesize,commandchars=\\\{\}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.pos = "H",
                      warning = FALSE,
                      message = FALSE)

```

## Contents
 - [Introduction](#introduction)
 - [Research question](#research-question)
 - [Packages and requirements](#packages-and-requirements)
 - [Preparation of the data](#preparation-of-the-data)
 - [Normalization and selection of the most variable features](#normalization-and-selection-of-the-most-variable-features)
 - [Dimension reduction with Principal Component Analysis](#dimension-reduction-with-principal-component-analysis)
 - [Clustering and UMAP visualisation](#clustering-and-umap-visualisation)
 - [Conclusion](Conclusion)
 - [Discussion](Discussion)
 - [Save the Seurat object as RDS file](#save-the-seurat-object-as-rds)
 - [Sources](#sources)

___
## Introduction
This RMarkdown describes an analysis of the single-nucleus RNA-sequencing count matrix from [GSE138852](https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE138852&format=file&file=GSE138852%5Fcounts%2Ecsv%2Egz) using Seurat. The dataset consists of a count matrix containing raw UMI counts derived from snRNAseq of human entorhinal cortex nuclei. Quality control of the nuclei was already performed according to the information provided in the article by Grubman _et al_., with low-quality cells removed during prior filtering steps.

The scope of this RMarkdown includes normalization of the count matrix, identification of highly variable features, and dimensionality reduction. Cell-type annotation is performed using BRETIGEA but is not covered within this RMarkdown.

## Research question
For this project the main research question is _"How can a reliable and reproducible analysis pipeline be set up to identify cell type-specific differential gene expression in Alzheimer's disease using a filtered expression matrix generated from snRNA-seq data?"_. Initially the research question was aimed at creating the entire pipeline with raw sequencing data as a starting point however, the publicly available data was lacking information on cellbarcodes and UMIs. Therefore the expression matrix published for GSE138852 was used as a starting point.

The sub-question(s) that is answered in this RMarkdown is as followed;

 - Can the most variable features be identified with the filtered expression count matrix as input? 
 - How should the count matrix be used, should the groups be separated (healthy vs. Alzheimers disease)? 
 
___
## Packages and Requirements
In order to run the analysis in Seurat a few packages need to be installed first. Packages and package versions are managed using ```renv```, instructions for setting up the R project with ```renv``` are given in the [README](https://github.com/ProjecticumDlerpDs/ad_transcriptomics_dge/blob/main/README.md) of the GitHub repository *ad_transcriptomics_dge*. 

```{r Load_packages, include=FALSE}
# Load the required packages
library(dplyr)
library(Seurat)
library(patchwork)
library(here)
library(ggplot2)
```

## Preparation of the data

### About the count matrix
The count matrix was downloaded from the GEO-database. The count matrix is a table which stores the gene expression measurements of all measured nuclei. Usually the matrix contains the raw expression measurements but in the case of GSE138852 some filters were already applied, the matrix contains;

 - 13214 nuclei (before filtering 14876 nuclei were measured)
 - Only nuclei with less then 10% of the total features being mitochondrial
 - None of the features that are related to postmortem interval
 - Only cells that are within the 5th - 95th percentile regarding gene and UMI count. 

Navigate to the directory *ad_transcriptomics/scripts* and run the script for downloading the count matrix from the terminal with the following command:
```{bash run_script_06_download_countmatrix.sh, eval = FALSE, echo = TRUE}
bash 06_download_countmatrix.sh
```

**Script 06_download_countmatrix.sh**
```{bash script_06_download_countmatrix.sh, eval = FALSE, echo = TRUE}
#!/bin/bash
# Script to download countmatrix from GEO-database

# Stop when errors occur
set -euo pipefail

# Path to directories 
project="$HOME/ad_transcriptomics_dge"
output="$project/raw_data"

# Create output directory
mkdir -p $output

# Download countmatrix from 
wget -P $output https://ftp.ncbi.nlm.nih.gov/geo/series/GSE138nnn/GSE138852/suppl/GSE138852_counts.csv.gz

echo "The count matrix is saved in: $output"
```

### Loading the dataset and creating a Seurat object
The count matrix was stored in a comma separated values file (.csv-file). To be able to use this file for creating a Seurat object ```read.csv()``` was used to create a data.frame first. Then the data.frame was used as input for the function ```CreateSeuratObject()```. 

```{r Load_data_set, include=FALSE}
# Load the GSE138852 countmatrix as a data.frame
count_matrix <- read.csv(file = here("raw_data", "GSE138852_counts.csv"),
  row.names = 1,
  check.names = FALSE
  )

# Initialize the Seurat object with the pre-filtered count matrix
count_matrix <- CreateSeuratObject(counts = count_matrix, project = "GSE138852")
```
\newpage

### Pre-processing of the dataset
Since the count matrix had already been pre-processed as described in "[About the count matrix](#About-the-count-matrix)", initially no additional pre-processing is performed in order to check if the dataset was indeed pre-processed as stated in the article by Grubman _et al_ [(1)](#sources). A column was added containing the percentage of mitochondrial features. A violin plot was generated and visualized in figure 1. 
After inspecting the unfiltered data in figure 1 quality-control filtering was applied to retain only nuclei with ≤5% mitochondrial RNA and ≥300 detected genes. The dataset was visualized in a violin plot after filtering in figure 2. To verify the quality of the dataset a scatterplot was created of nCount_RNA vs Percent_mt and nCount_RNA vs nFeature_RNA (figure 3). Here it is shown that cells with a lower RNA count show a higher percentage of mitochondrial RNA and that cells that show a high amount of RNA counts also show a higher amount of unique genes.

```{r Preprocessing_of_data_set}
# Add a column containing the percentage of genes starting with MT (mitochondrial).
count_matrix[["Percent_mt"]] <- PercentageFeatureSet(count_matrix, pattern = "^MT-")
```

```{r Visualize_dataframe_without_filters, fig.cap="Violin plots of quality-control metrics. Shown are the number of detected genes per nucleus (nFeature_RNA), total UMI counts per nucleus (nCount_RNA), and the percentage of mitochondrial gene expression (Percent_mt). The distributions of these plots are as expected with prior quality-control filtering as described by Grubman et al.", fig.pos='!htbp'}
# Visualize dataframe as a violin plot
VlnPlot(count_matrix, features = c("nFeature_RNA", "nCount_RNA", "Percent_mt"), ncol = 3)+ 
  NoLegend()
```

```{r Adjust_filtering_of_data_set}
count_matrix <- subset(count_matrix, subset = nFeature_RNA > 300 & Percent_mt < 5)
```

```{r Visualize_dataframe, fig.cap="Violin plots of quality-control metrics. Shown are the number of detected genes per nucleus (nFeature_RNA), total UMI counts per nucleus (nCount_RNA), and the percentage of mitochondrial gene expression (Percent_mt). Nuclei were retained for downstream analyses if nFeature_RNA was at least 300 detected genes and percentage mitochondrial RNA was no more then 5%, filtering out low-quality nuclei.", fig.pos='!htbp'}
# Visualize dataframe as a violin plot
VlnPlot(count_matrix, features = c("nFeature_RNA", "nCount_RNA", "Percent_mt"), ncol = 3)+ 
  NoLegend()
```

```{r Visualize_dataframe_as_scatterplot, fig.cap="Scatterplot to visualise a relationship between the percentage of mitochondrial RNA versus the total amount of RNA and to visualize the relationship between the total amount of features and the total amount of RNA", fig.pos='!htbp'}

# Visualize dataframe in scatterplots
plot1 <- FeatureScatter(count_matrix,
                        feature1 = "nCount_RNA",
                        feature2 = "Percent_mt",
                        plot.cor = FALSE) +
  ggtitle("Mitochondrial percentage\nverus RNA counts") +
  xlab("Total RNA counts") +
  ylab("Percentage MT genes") +
  theme_classic()+ 
  NoLegend()
plot2 <- FeatureScatter(count_matrix, 
                        feature1 = "nCount_RNA",
                        feature2 = "nFeature_RNA",
                        plot.cor = FALSE) +
  ggtitle("RNA features\nversus RNA counts") +
  xlab("Total RNA counts") +
  ylab("RNA features") +
  theme_classic()+ 
  NoLegend()
plot1 + plot2
```

___

## Normalization and selection of the most variable features

### Normalization of the dataset
The count matrix was normalized using ```NormalizeData()```. For normalization the default settings of Seurat were used which means that the argument ```LogNormalize``` was used. With this argument feature counts per cell are divided by the total counts and multiplied by the ```scale.factor```. The default setting for ```scale.factor``` is 10.000.

```{r Normalize_data, include=FALSE}
# To normalize the data count_matrix is replaced by a new object for count_matrix in which the data is normalized by using NormalizeData with the default settings LogNormalize and scale.factor=10000.
count_matrix <- NormalizeData(count_matrix) 
```

### Identification of highly variable features
After normalizing the dataset ```FindVariableFeatures()``` was used to detect the most variable features. The selection method that was used is ```vst```, and ```nFeatures``` was set at 2000. A scatter plot of the output of ```FindVariableFeatures()``` generated and visualized in figure 4. 

```{r Find_variable_features, fig.cap="Scatterplot of highly variable features in the GSE138852 count matrix. Average gene expression is plotted on the x-axis against standardized variance on the y-axis. Highly variable genes (red dots) were selected for further analysis. The top 20 most variable genes were labeled.", fig.pos='!htbp', fig.height=6}
# Find variable features 
count_matrix <- FindVariableFeatures(count_matrix, selection.method = "vst", nfeatures = 2000)
# Identify the 20 most highly variable genes
top20 <- head(VariableFeatures(count_matrix), 20)
# plot variable features with labels (with LabelPoints())
plot1 <- VariableFeaturePlot(count_matrix) %>%
  LabelPoints(points = top20, repel = TRUE, size = 3) +
  labs(
    title = "Features of GSE138852 dataset",
    subtitle = "Top 20 most variable features are labeled",
    x = "Average expression",
    y = "Standardized variance"
  ) +
  theme_classic() +
  theme(
    text = element_text(size = 10),
    plot.title = element_text(size = 11),
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 9),
    legend.position = "bottom"
  ) 
plot1
```


## Dimension reduction with Principal Component Analysis
### Data scaling
To prepare the dataset for Principal Component Analysis (PCA), scaling was performed with ```ScaleData```. 

```{r Scaling_data}
# Scale the data of the count matrix 
count_matrix <- ScaleData(count_matrix)
```

### Linear Dimension Reduction
The scaled data is used as input for Principal Component Analysis (PCA). PCA is performed with the function ```RunPCA()```, within this function "```npcs```" was kept at 50 which is the default setting to determine how many PCs should be computed.

```{r Run_PCA}
# Run PCA with RunPCA() on the count_matrix
count_matrix <- RunPCA(count_matrix, features = VariableFeatures(object = count_matrix))
```
After running PCA the first 20 PCs were visualized in an elbow plot (figure 5) with the standard deviation on the y-axis and the different PCs in ranked order on the x-axis. This plot was used to determine how many PCs should be retained for further analysis by identifying the elbow. The elbow is the point where the decrease in standard deviation between PCs is minimizing. Beyond this point, remaining variation is more likely to be caused by noise rather than biologically meaningful variance. In figure 5 the elbow was located around PC8, to be certain that no biologically meaningful variance was lost PC1 throughout PC9 for further analysis. 

```{r Elbowplot_PCA, fig.height=3, fig.cap="Elbow plot of principal components. This plot shows the standard deviation of each principal component. Around PC8 an elbow is observed after PC8, nFeatures = 2000."}
# Create an elbowplot of count_matrix after PCA
ElbowPlot(count_matrix)
```

To visualize which genes contribute to the different principal components, PCA loading plots were generated for PC1 through PC3 (Figure 6). These plots highlight the genes with the strongest contributions to each component. This way the source of variation within a PC can be interpreted. By comparing the top-loading genes with known marker genes for brain tissue specific cells it was assessed whether the variation in each principal is biologically meaningful. For example in PC1 and PC2 the top-loading genes where mainly found to be synaptic genes which could be found in neurones, such as  NRXN1 [(2)](#sources) and SCN1A [(3)](#sources). While PC3 is dominated by immune (microglial) genes such as HLA-DRA, CD86 and PTPRC [(4)](#sources). 

```{r Plot_PCA1, fig.height=10, fig.width=7, fig.cap="Principal component loading plot showing the top contributing genes for PC1–PC3, based on 2000 highly variable genes derived from the filtered snRNA-seq expression matrix."}
# Visualize PC1 through PC3 in a PC feature loading plot for count_matrix
VizDimLoadings(count_matrix, dims = 1:3, reduction = "pca")
``` 

\newpage

## Clustering and UMAP visualisation

### Find nearest neighbours
After PCA the data was reduced to 9 principal components, these components were used as input for clustering. Clustering was done with ```FindNeighbours()```, the dimensions were set at 1:9 and ```k.param``` was kept at the default value of 20 since lower values (k.param=10) caused the clusters scatter more and higher values (k.param=30) didn't seem to improve the data. 
The output of ```FindNeighbours()``` was used as input for ```RunUMAP``` and for ```FindClusters()```. ```FindClusters()``` was used to verify which cells have a expression pattern more similar to each other and was eventually used to see to which cell type the different cells belong. The clustering resolution was tested at multiple values (0.5, 0.7, and 1.0) in order to determine the resolution that produced the most stable clusters without losing the specificity to annotate cell types downstream in analysis. The clusters with the different resolutions were shown in three UMAP plots in figure 7, figure 8 and figure 9.

```{r Plot_UMAP1, fig.height=5, fig.width=6, fig.cap= "UMAP visualization of GSE138852. Cells are visualized in two-dimensional UMAP space using the first 9PCs. Each point represents a single nucleus, coloured by cluster assignment identified at resolution 0.5."}
# FindNeighbours of PC1-9, with k.param = 20.
count_matrix <- FindNeighbors(count_matrix, dims = 1:9)
# RunUMAP with input of find neighbours and PCA to determine coordinates of all cells in UMAP space
count_matrix <- RunUMAP(count_matrix, dims = 1:9)  
# FindClusters with different resolutions
count_matrix1 <- FindClusters(count_matrix, resolution = 0.5, verbose=FALSE)
count_matrix2 <- FindClusters(count_matrix, resolution = 0.7, verbose=FALSE)
count_matrix3 <- FindClusters(count_matrix, resolution = 1.0, verbose=FALSE)
# Plot the UMAP of res 0.5
DimPlot(count_matrix1, reduction = "umap")
```

```{r Plot_UMAP2, fig.height=5, fig.width=6, fig.cap= "UMAP visualization of GSE138852. Cells are visualized in two-dimensional UMAP space using the first 9PCs. Each point represents a single nucleus, coloured by cluster assignment identified at resolution 0.7."}
# Plot the UMAP of res 0.7
DimPlot(count_matrix2, reduction = "umap")
```

```{r Plot_UMAP3, fig.height=5, fig.width=6, fig.cap= "UMAP visualization of GSE138852. Cells are visualized in two-dimensional UMAP space using the first 9PCs. Each point represents a single nucleus, coloured by cluster assignment identified at resolution 1.0."}
# Plot the UMAP of res 1.0
DimPlot(count_matrix3, reduction = "umap")
```

\newpage

## Conclusion
In order to check if the count matrix contained the data as described by the article by Grubman _et al_. and the GEO-database (Accession number: GSE138852) the data was plotted in three violin plots (figure 1). These violin plots confirm that the filtering was applied as stated in the article by Grubman et al., most nuclei show between 400 to 800 detected features, the amount of total RNA molecules detected is between 500 - 1000 for most nuclei and the percentage of mitochondrial features is around 1,5% with no values >10%. 
After applying filters for percentage of mitochondrial genes and the unique number of genes detected, the data was plotted in figure 3. The quality control plots in figure 3 indicate that the cells retained in this dataset are of high quality. The percentage of mitochondrial gene expression is low for the majority of nuclei and shows the expected relationship with total RNA counts (lower counts usually means higher percentage of mitochondrial genes). The relationship between the amount of unique features and the amount of RNA counts show a strong correlation, this is also an expected pattern since ,when more UMIs are captured from a cell, more genes are usually detected this cell as well. 

The elbow plot in figure 5 shows that the most variation in the dataset is caused by PC1 through PC8, to ensure no relevant data is lost PC1 through PC9 were retained for downstream analysis. In figure 6 the PCA loading were plotted, the PCA loadings of PC1 through PC3 indicate that some of the contributing genes are known as neuronal and microglial genes. It can be concluded that the amount of features used in PCA and the amount of PCs calculated are sufficient to capture relevant biological variation within the GSE138852 dataset.

The UMAP plots in figure 8, with resolution 0.7, shows the most stable, well defined, clusters. Therefore, this resolution is used for downstream analysis.

In relation to the research question the following is concluded;
 
 - Can the most variable features be identified with the filtered expression count matrix as input? 

The most variable features can and were identified with the filtered expression matrix as input.
 
 - How should the count matrix be used, should the groups be separated (healthy vs. Alzheimers disease)? 
 
For this part of analysis the data was not separated. The dataset was kept as one to identify the different cell types based on overall gene expression data. This is done since for differential gene expression analysis the comparison should be made between healthy and diseased condition in the same group of cells. This makes it crucial that cell types are defined consistently across both conditions.

## Discussion
This part of analysis was performed with a filtered expression matrix as input. Using pre-filtered data makes analysis a lot faster, however there is no way of knowing the difference between the dataset pre-filtering and post-filtering. If this RMarkdown would be used on a different dataset the same functions could be used but processing and of the data should be re-evaluated. 

The different settings that were used such as nFeatures, number of PCs and settings for resolution seem to produce relevant PCs and stable clusters in UMAP visualization. However no cell types were identified yet, this is to be done in **03_celltype_annotation_bretigea.Rmd**. This is a critical next step to validate whether the identified clusters correspond to known cell types. Confirmation of the reliability of the chosen settings will depend on the expression of marker genes within the clusters.

### Save the Seurat object as RDS
After analysis the Seurat object with the correct parameters was saved as .RDS file. The resolution of the object was stored in the title.

```{r Save_seurat_object_to_RDS, message=TRUE}
# Create the output directory with here, show no warnings if directory allready exists
dir.create(here("analyses","seurat"), showWarnings = FALSE)

# Define output directory
output_dir <- here("analyses", "seurat")

# Define output file
output_file <- here(output_dir, "seurat_umap_clustering_res0.7.rds")

# Save the Seurat object after analysis as RDS object to be available for further analysis
saveRDS(
  count_matrix2,
  file = output_file
)

# Print message to inform about the name of the file and the directory
message(
  "The Seurat object was stored as:\n",
  "  File: ", basename(output_file), "\n",
  "  Directory: ", output_dir
)
```


## Sources
 
 - 1: Grubman, A., Chew, G., Ouyang, J. F., Sun, G., Choo, X. Y., McLean, C., Simmons, R. K., Buckberry, S., Vargas-Landin, D. B., Poppe, D., Pflueger, J., Lister, R., Rackham, O. J. L., Petretto, E., & Polo, J. M. (2019). A single-cell atlas of entorhinal cortex from individuals with Alzheimer’s disease reveals cell-type-specific gene expression regulation. Nature Neuroscience, 22(12), 2087–2097.
 - 2: [https://www.cell.com/cell/fulltext/S0092-8674(17)31246-1](https://www.cell.com/cell/fulltext/S0092-8674(17)31246-1), Accessed on 03 January 2026.
 - 3: [https://pmc.ncbi.nlm.nih.gov/articles/PMC12559145/](https://pmc.ncbi.nlm.nih.gov/articles/PMC12559145/), Accessed on 12 January 2026.
 - 4: Jurga, A. M., Paleczna, M., & Kuter, K. Z. (2020). Overview of general and discriminating markers of differential microglia phenotypes. Frontiers in Cellular Neuroscience, 14, 198.
 
 

