---
title: "Identification most variable features in Seurat"
author: "Meryem Stroosma"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    latex_engine: xelatex
    fig_caption: true

# Use colours for hyperlinks
colorlinks: true
linkcolor: purple
urlcolor: purple
citecolor: purple

header-includes:
  - \usepackage{fontspec}
  - \usepackage{fancyvrb}
  - \usepackage[section]{placeins}
  - \usepackage{float} 
  - \usepackage{blindtext}
  - \setmainfont{DejaVu Sans Condensed}
  - \setsansfont{DejaVu Sans}
  - \setmonofont{DejaVu Sans Mono}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{fontsize=\footnotesize,commandchars=\\\{\}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      fig.pos = "H")

```

## Contents
 - [Introduction](#introduction)
 - [Research question](#research-question)
 - [Packages and requirements](#packages-and-requirements)
 - [Preparation of the data](#preparation-of-the-data)
 - [Normalization and selection of the most variable features](#normalization-and-selection-of-the-most-variable-features)
 - [Dimension reduction with Principal Component Analysis](#dimension-reduction-with-principal-component-analysis)
 - [Conclusion](Conclusion)
 - [Discussion](Discussion)

___
## Introduction
This RMarkdown describes an analysis of the single-nucleus RNA-sequencing count matrix from [GSE138852](https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE138852&format=file&file=GSE138852%5Fcounts%2Ecsv%2Egz) using Seurat. The dataset consists of a count matrix containing raw UMI counts derived from snRNAseq of human entorhinal cortex nuclei. Quality control of the nuclei was already performed according to the information provided in the article by Grubman _et al_., with low-quality cells removed during prior filtering steps.

The scope of this RMarkdown includes normalization of the count matrix, identification of highly variable features, and dimensionality reduction. Cell-type annotation is performed using BRETIGEA but is not covered within this RMarkdown.

## Research question
For this project the main research question is _"How can a reliable and reproducible analysis pipeline be set up to identify cell type-specific differential gene expression in Alzheimer's disease using a filtered expression matrix generated from snRNA-seq data?"_. Initially the research question was aimed at creating the entire pipeline with raw sequencing data as a starting point however, the publicly available data was lacking information on cellbarcodes and UMIs. Therefore the expression matrix published for GSE138852 was used as a starting point.

The sub-question(s) that is answered in this RMarkdown is as followed;

 - Can the most variable features be identified with the filtered expression count matrix as input? 
 - How should the count matrix be used, should the groups be separated (healthy vs. Alzheimers disease)? 
 
___
## Packages and Requirements
In order to run the analysis in Seurat a few packages need to be installed first. Packages and package versions are managed using ```renv```, instructions for setting up the R project with ```renv``` are given in the [README](https://github.com/ProjecticumDlerpDs/ad_transcriptomics_dge/blob/main/README.md) of the GitHub repository *ad_transcriptomics_dge*. 

```{r Load_packages, include=FALSE}
# Load the required packages
library(dplyr)
library(Seurat)
library(patchwork)
library(here)
library(ggplot2)
```

## Preparation of the data

### About the count matrix
The count matrix was downloaded from the GEO-database. The count matrix is a table which stores the gene expression measurements of all measured nuclei. Usually the matrix contains the raw expression measurements but in the case of GSE138852 some filters were already applied, the matrix contains;

 - 13214 nuclei (before filtering 14876 nuclei were measured)
 - Only nuclei with less then 10% of the total features being mitochondrial
 - None of the features that are related to postmortem interval
 - Only cells that are within the 5th - 95th percentile regarding gene and UMI count. 

Navigate to the directory *ad_transcriptomics/scripts* and run the script for downloading the count matrix from the terminal with the following command:
```{bash run script 06_download_countmatrix.sh, eval = FALSE, echo = TRUE}
bash 06_download_countmatrix.sh
```

### Loading the dataset and creating a Seurat object
The count matrix was stored in a comma separated values file (.csv-file). To be able to use this file for creating a Seurat object ```read.csv()``` was used to create a data.frame first. Then the data.frame was used as input for the function ```CreateSeuratObject()```. 

```{r Load_data_set, include=FALSE}
# Load the GSE138852 countmatrix as a data.frame
count_matrix <- read.csv(file = here("raw_data", "GSE138852_counts.csv"),
  row.names = 1,
  check.names = FALSE
  )

# Initialize the Seurat object with the pre-filtered count matrix
count_matrix <- CreateSeuratObject(counts = count_matrix, project = "GSE138852")
```
\newpage

### Pre-processing of the dataset
Since the count matrix had already been pre-processed as described in "[About the count matrix](#About-the-count-matrix)", no additional pre-processing is performed at this point. However to check if the dataset was indeed pre-processed as stated in the article by Grubman _et al_. a column was added containing the percentage of mitochondrial features. A violin plot was generated and visualized in figure 1. 

```{r Preprocessing_of_data_set}
# Add a column containing the percentage of genes starting with MT (mitochondrial).
count_matrix[["Percent_mt"]] <- PercentageFeatureSet(count_matrix, pattern = "^MT-")
```

```{r Visualize_dataframe, warning=FALSE, fig.cap="Violin plots of quality-control metrics. Shown are the number of detected genes per nucleus (nFeature_RNA), total UMI counts per nucleus (nCount_RNA), and the percentage of mitochondrial gene expression (Percent_mt). The distributions of these plots are as expected with prior quality-control filtering as described by Grubman et al.", fig.pos='!htbp'}
# Visualize dataframe as a violin plot
VlnPlot(count_matrix, features = c("nFeature_RNA", "nCount_RNA", "Percent_mt"), ncol = 3)
```

___
## Normalization and selection of the most variable features
### Normalization of the dataset
The count matrix was normalized using ```NormalizeData()```. For normalization the default settings of Seurat were used which means that the argument ```LogNormalize``` was used. With this argument feature counts per cell are divided by the total counts and multiplied by the ```scale.factor```. The default setting for ```scale.factor``` is 10.000.

```{r Normalize_data, include=FALSE, warning=FALSE}
# To normalize the data count_matrix is replaced by a new object for count_matrix in which the data is normalized by using NormalizeData with the default settings LogNormalize and scale.factor=10000.
count_matrix <- NormalizeData(count_matrix) 
```

### Identification of highly variable features
After normalizing the dataset ```FindVariableFeatures()``` was used to detect the most variable features. The selection method that was used is ```vst```, and ```nFeatures``` was set at 2000. A scatter plot of the output of ```FindVariableFeatures()``` generated and visualized in figure 2. 

```{r Find_variable_features, warning=FALSE, message=FALSE, fig.cap="Scatterplot of highly variable features in the GSE138852 count matrix. Average gene expression is plotted on the x-axis against standardized variance on the y-axis. Highly variable genes (red dots) were selected for further analysis. The top 10 most variable genes were labeled.", fig.pos='!htbp', fig.height=7}
# Find variable features 
count_matrix <- FindVariableFeatures(count_matrix, selection.method = "vst", nfeatures = 2000)
# Identify the 20 most highly variable genes
top20 <- head(VariableFeatures(count_matrix), 20)
# plot variable features with labels (with LabelPoints())
plot1 <- VariableFeaturePlot(count_matrix) %>%
  LabelPoints(points = top20, repel = TRUE, size = 3) +
  labs(
    title = "Features of GSE138852 dataset",
    subtitle = "Top 20 most variable features are labeled",
    x = "Average expression",
    y = "Standardized variance"
  ) +
  theme_classic() +
  theme(
    text = element_text(size = 10),
    plot.title = element_text(size = 11),
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 9),
    legend.position = "bottom"
  ) 
plot1
```

## Dimension reduction with Principal Component Analysis
### Data scaling
To prepare the dataset for Principal Component Analysis (PCA), scaling was performed with ```ScaleData```. 

```{r Scaling_data, warning=FALSE, message=FALSE}
# Scale the data of the count matrix 
count_matrix <- ScaleData(count_matrix)
```

### Linear Dimension Reduction
The scaled data is used as input for Principal Component Analysis (PCA). PCA is performed with the function ```RunPCA()```, within this function "```npcs```" was kept at 50 which is the default setting to determine how many PCs should be computed.

```{r Run_PCA, warning=FALSE, message=FALSE}
# Run PCA with RunPCA() on the count_matrix
count_matrix <- RunPCA(count_matrix, features = VariableFeatures(object = count_matrix))
```
After running PCA the first 20 PCs were visualized in an elbow plot (figure 3) with the standard deviation on the y-axis and the different PCs in ranked order on the x-axis. This plot was used to determine how many PCs should be retained for further analysis by identifying the elbow. The elbow is the point where the decrease in standard deviation between PCs is minimizing. Beyond this point, remaining variation is more likely to be caused by noise rather than biologically meaningful variance. In figure 3 the elbow was located around PC8, to be certain that no biologically meaningful variance was lost PC1 throughout PC9 for further analysis. 

```{r Elbowplot_PCA1, warning=FALSE, message=FALSE, fig.height=3, fig.cap="Elbow plot of principal components. This plot shows the standard deviation of each principal component. Around PC8 an elbow is observed after PC8, nFeatures = 2000."}
# Create an elbowplot of count_matrix after PCA
ElbowPlot(count_matrix)
```

To visualize which genes contribute to the different principal components, a PCA loading plot was created.  

```{r Plot_PCA1, warning=FALSE, message=FALSE, fig.height=10, fig.cap="Principal component loading plot showing the top contributing genes for PC1–PC4, based on 2000 highly variable genes derived from the filtered snRNA-seq expression matrix."}
# Visualize PC1 through PC4 in a PC feature loading plot for count_matrix
VizDimLoadings(count_matrix, dims = 1:4, reduction = "pca")
``` 

Explain how the resolution was chosen, at 0.5 the clusters seem clean and compact without too much subclustering. The next step is to see if different celltypes can be annotated based on these clusters, 0.7 also seems an okay resolution however the larger clusters are splitting into subclusters which are less homogenous 

```{r Cluster_cells, include=FALSE}
# Find Nearest Neighbors for dataset, default setting was used for k.param. 9 PCs were used. 
count_matrix <- FindNeighbors(count_matrix, dims = 1:9)
# Identification of clusters, resolution was set at 0.5. 
count_matrix1 <- FindClusters(count_matrix, resolution = 0.5)
# Identification of clusters, resolution was set at 0.7. 
count_matrix2 <- FindClusters(count_matrix, resolution = 0.7)
# Identification of clusters, resolution was set at 1.0. 
count_matrix3 <- FindClusters(count_matrix, resolution = 1.0)
```

```{r Plot_UMAP1, warning=FALSE, message=FALSE, fig.cap= "UMAP of object with nfeatures=2000. Cells are visualised in two-dimensional UMAP space based on the first 9 PCs. Each point represents a single nucleus and colours indicate clusters identified using FindClusters (resolution = 0.5)."}
# Create a UMAP visualisation after finding clusters for PCs 1 through 9
count_matrix1 <- RunUMAP(count_matrix1, dims = 1:9)

# Plot the UMAP
DimPlot(count_matrix1, reduction = "umap")
```

```{r Plot_UMAP2, warning=FALSE, message=FALSE, fig.cap= "UMAP of object with nfeatures=2000. Cells are visualised in two-dimensional UMAP space based on the first 9 PCs. Each point represents a single nucleus and colours indicate clusters identified using FindClusters (resolution = 0.7)."}
# Create a UMAP visualisation after finding clusters for PCs 1 through 9
count_matrix2 <- RunUMAP(count_matrix2, dims = 1:9)

# Plot the UMAP
DimPlot(count_matrix2, reduction = "umap")
```

```{r Plot_UMAP3, warning=FALSE, message=FALSE, fig.cap= "UMAP of object with nfeatures=2000. Cells are visualised in two-dimensional UMAP space based on the first 9 PCs. Each point represents a single nucleus and colours indicate clusters identified using FindClusters (resolution = 1.0)."}
# Create a UMAP visualisation after finding clusters for PCs 1 through 9
count_matrix3 <- RunUMAP(count_matrix3, dims = 1:9)

# Plot the UMAP
DimPlot(count_matrix3, reduction = "umap")
```


```{r Save_seurat_object_to_RDS}
# Create the output directory with here, show no warnings if directory allready exists
dir.create(here("analyses","seurat"), showWarnings = FALSE)

# Save the seurat object after analysis as RDS object to be available for furthur analysis
saveRDS(
  count_matrix1,
  file = here("analyses", "seurat", "seurat_umap_clustering_res0.5.rds")
)
```



## Conclusion
In order to check if the count matrix contained the data as described by the article by Grubman _et al_. and the GEO-database (Accession number: GSE138852) the data was plotted in three violin plots (figure 1). These violin plots confirm that the filtering was applied as stated in the article by Grubman et al., most nuclei show between 400 to 800 detected features, the amount of total RNA molecules detected is between 500 - 1000 for most nuclei and the percentage of mitochondrial features is around 1,5% with no values >10%. 

