---
title: "Celltype_annotation_SingleR"
author: "Meryem Stroosma"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    latex_engine: xelatex
    fig_caption: true
    
params:
  resolution: "0.5"
  dataset_name: "GSE138852"

# Use colours for hyperlinks
colorlinks: true
linkcolor: purple
urlcolor: purple
citecolor: purple

header-includes:
  - \usepackage{fontspec}
  - \usepackage{fancyvrb}
  - \usepackage[section]{placeins}
  - \usepackage{float} 
  - \usepackage{blindtext}
  - \setmainfont{DejaVu Sans Condensed}
  - \setsansfont{DejaVu Sans}
  - \setmonofont{DejaVu Sans Mono}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{fontsize=\footnotesize,commandchars=\\\{\}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      error = FALSE)
```

## Contents
 - [Introduction](#introduction)
 - [Research question](#research-question)
 - [Packages and Requirements](#packages-and-requirements)
 
## Introduction


## Research question


 ___
## Packages and Requirements
In order to run the analysis in Seurat a few packages need to be installed first. Packages and package versions are managed using ```renv```, instructions for setting up the R project with ```renv``` are given in the [README](https://github.com/ProjecticumDlerpDs/ad_transcriptomics_dge/blob/main/README.md) of the GitHub repository *ad_transcriptomics_dge*. 

```{r setup_libraries, echo=FALSE, include=FALSE}
# Load the required packages
library(Seurat)
library(here)
library(SingleR)
library(scRNAseq)
library(SummarizedExperiment)
```

```{r setup_find_seuratobject_function, echo=FALSE}
# A function was created to load a Seurat object created with any given resolution, the resolution can be given by the user upon knitting this RMarkdown. 


load_seurat_by_resolution <- function(resolution,
                                      dir = c("analyses", "seurat"),
                                      prefix = "seurat_umap_clustering_") {
# Create a pattern to search for a filename, the clustering resolution can be given using params. The default resolution is 0.5.
  res_pattern <- paste0("res", gsub("\\.", "\\\\.", resolution), "\\.rds$")

# Search in the specified directory for Seurat objects matching the pattern.
  files <- list.files(
    path = do.call(here, as.list(dir)),
    pattern = paste0("^", prefix, res_pattern),
    full.names = TRUE
  )
# If no files are found then stop execution.
  if (length(files) == 0)
    stop("No Seurat object found for resolution ", resolution)

# If multiple files with the same pattern are found then stop execution.
  if (length(files) > 1)
    stop("Multiple Seurat objects found for resolution ", resolution)

# Print which file is used to load the Seurat object to be used for further analysis.
  message("Loading Seurat object: ", basename(files))
  readRDS(files)
}
```


```{r Load_Seurat_object}
# Load the seurat object based on the param that was used for resolution
seurat_obj <- load_seurat_by_resolution(params$resolution)
```

```{r Extract_raw_counts}
normalised_counts <- LayerData(seurat_obj, assay = "RNA", layer = 'data')

```

```{r Get_reference_dataset}

ref <- DarmanisBrainData()

# If SingleR expects logcounts, create it from counts without scater
if (!"logcounts" %in% assayNames(ref)){ 
  
# counts(ref) should exist for DarmanisBrainData; use SummarizedExperiment::assay to be explicit
cts <- SummarizedExperiment::assay(ref, "counts")

# log1p works on sparse dgCMatrix too
SummarizedExperiment::assay(ref, "logcounts") <- log1p(cts)
}
```

```{r Run_SingleR}
labels <- colData(ref)[["cell.type"]]
keep <- !is.na(labels)

celltype_annotation <- SingleR(
  test   = normalised_counts,
  ref    = ref[, keep],
  labels = labels[keep],
  assay.type.ref = "logcounts"
)

```

```{r Plot_heatmap}
plotScoreHeatmap(celltype_annotation)
```


