---
title: "Celltype annotation with BRETIGEA"
author: "Meryem Stroosma"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    latex_engine: xelatex
    fig_caption: true
    
params:
  resolution: "0.5"
  dataset_name: "GSE138852"

# Use colours for hyperlinks
colorlinks: true
linkcolor: purple
urlcolor: purple
citecolor: purple

header-includes:
  - \usepackage{fontspec}
  - \usepackage{fancyvrb}
  - \usepackage[section]{placeins}
  - \usepackage{float} 
  - \usepackage{blindtext}
  - \setmainfont{DejaVu Sans Condensed}
  - \setsansfont{DejaVu Sans}
  - \setmonofont{DejaVu Sans Mono}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{fontsize=\footnotesize,commandchars=\\\{\}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      error = FALSE)
```
## Contents 
 - [Introduction](#introduction) 
 - [Research question](#research-question) 
 - [Packages and requirements](#packages-and-requirements)
 - [Load Seurat object](#load-seurat-object)
 - [Marker discovery per cluster](#marker-discovery-per-cluster)
 - [Marker plots](#marker-plots)
 - [Cell type annotation with BRETIGEA](#cell-type-annotation-with-bretigea)
 - [Assign labels back to Seurat](#assign-labels-back-to-seurat)
 - [UMAP visualization](#umap-visualization)
 - [Save outputs](#save-outputs)

___ 

## Packages and Requirements 
In order to run the analysis in Seurat a few packages need to be installed first. Packages and package versions are managed using ```renv```, instructions for setting up the R project with ```renv``` are given in the [README](https://github.com/ProjecticumDlerpDs/ad_transcriptomics_dge/blob/main/README.md) of the GitHub repository *ad_transcriptomics_dge*.

```{r setup_libraries, echo=FALSE, include=FALSE}
# Load the required packages
library(dplyr)
library(Seurat)
library(BRETIGEA)
library(patchwork)
library(here)
library(ggplot2)
```

```{r setup_find_seuratobject_function, echo=FALSE}
# A function was created to load a Seurat object created with any given resolution, the resolution can be given by the user upon knitting this RMarkdown. 


load_seurat_by_resolution <- function(resolution,
                                      dir = c("analyses", "seurat"),
                                      prefix = "seurat_umap_clustering_") {
# Create a pattern to search for a filename, the clustering resolution can be given using params. The default resolution is 0.5.
  res_pattern <- paste0("res", gsub("\\.", "\\\\.", resolution), "\\.rds$")

# Search in the specified directory for Seurat objects matching the pattern.
  files <- list.files(
    path = do.call(here, as.list(dir)),
    pattern = paste0("^", prefix, res_pattern),
    full.names = TRUE
  )
# If no files are found then stop execution.
  if (length(files) == 0)
    stop("No Seurat object found for resolution ", resolution)

# If multiple files with the same pattern are found then stop execution.
  if (length(files) > 1)
    stop("Multiple Seurat objects found for resolution ", resolution)

# Print which file is used to load the Seurat object to be used for further analysis.
  message("Loading Seurat object: ", basename(files))
  readRDS(files)
}
```

## Load Seurat object
```{r Load_Seurat_object}
seurat_obj <- load_seurat_by_resolution(params$resolution)

# Ensure identities are clusters
Idents(seurat_obj) <- "seurat_clusters"
```

## Marker discovery per cluster

```{r Find_cluster_markers}
markers_all <- FindAllMarkers(
  seurat_obj,
  only.pos = TRUE,          # annotation-focused: positive markers only
  test.use = "wilcox",
  min.pct = 0.25,
  logfc.threshold = 0.25
)

top_markers <- markers_all %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 10) %>%
  ungroup()

top_markers
```

## Marker plots

### Heatmap of top markers

```{r Heatmap_top_markers, fig.height=8, fig.width=8}
top5 <- markers_all %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 5) %>%
  pull(gene) %>%
  unique()

# Scale only the genes we plot (keeps it fast)
seurat_obj <- ScaleData(seurat_obj, features = top5, verbose = FALSE)

DoHeatmap(seurat_obj, features = top5, group.by = "seurat_clusters") + NoLegend()
```

### DotPlot of canonical brain markers

```{r DotPlot_canonical_markers, fig.height=6, fig.width=10}
canonical <- c(
  "RBFOX3","SNAP25","SYT1",          # neurons
  "SLC17A7",                         # excitatory
  "GAD1","GAD2",                     # inhibitory
  "AQP4","ALDH1A1","SLC1A2","SLC1A3",# astrocytes
  "MBP","PLP1","MOG",                # oligodendrocytes
  "PDGFRA","CSPG4","VCAN",           # OPC
  "P2RY12","CX3CR1","C1QA",          # microglia
  "PECAM1","VWF","KDR",              # endothelial
  "PDGFRB","RGS5","ACTA2"            # pericyte/VSMC
)

canonical_present <- canonical[canonical %in% rownames(seurat_obj)]

DotPlot(seurat_obj, features = canonical_present) + RotatedAxis()
```

### FeaturePlot (UMAP) for key markers

```{r FeaturePlot_key_markers, fig.height=7, fig.width=10}
key_markers <- c("SLC17A7","GAD1","AQP4","MBP","P2RY12","PECAM1")
key_present <- key_markers[key_markers %in% rownames(seurat_obj)]

FeaturePlot(seurat_obj, features = key_present, ncol = 3)
```

## Cell type annotation with BRETIGEA

### Create expression matrix per cluster (average expression)
```{r create_expression_matrix_per_cluster}
pb <- AverageExpression(
  seurat_obj,
  assays = "RNA",
  layer  = "data"
)$RNA

# FIX: remove the 'g' prefix added by Seurat for identities starting with numbers
colnames(pb) <- sub("^g", "", colnames(pb))
```

### Select BRETIGEA marker reference and run scoring

```{r select_Bretigea_reference}
markers <- BRETIGEA::markers_df_human_brain
genes_use <- intersect(unique(markers$marker), rownames(pb))
pb_use <- as.matrix(pb[genes_use, , drop = FALSE])
```

```{r run_bretigea_on_clusters}
ct_res <- BRETIGEA::findCells(
  inputMat = pb_use,
  markers  = markers,
  nMarker  = 20,
  method   = "PCA",
  scale    = FALSE
)
```

### Create one label per cluster (max score)
```{r Create_label_per_cluster}
scores <- as.matrix(ct_res)

# clusters are rows; cell types are columns
cluster_to_celltype <- apply(scores, 1, function(x) colnames(scores)[which.max(x)])

# Store scores + label mapping for transparency
bretigea_cluster_scores <- as.data.frame(scores)
bretigea_cluster_scores$cluster <- rownames(scores)
bretigea_cluster_scores$BRETIGEA_label <- unname(cluster_to_celltype[rownames(scores)])

bretigea_cluster_scores
```

## Assign labels back to Seurat

```{r Return_labels_to_Seuratobj}
clusters <- as.character(Idents(seurat_obj))
celltype_by_cell <- unname(cluster_to_celltype[clusters])

celltype_by_cell[is.na(celltype_by_cell)] <- "Unknown"
names(celltype_by_cell) <- colnames(seurat_obj)

# Store BRETIGEA call as a metadata column
seurat_obj <- AddMetaData(seurat_obj, celltype_by_cell, col.name = "BRETIGEA_celltype")

# Optional: a more generic level-1 column (you can later edit/merge labels if needed)
seurat_obj$celltype_level1 <- seurat_obj$BRETIGEA_celltype

table(seurat_obj$BRETIGEA_celltype, useNA = "ifany")
```

## UMAP visualization

```{r Plot_celltypes_umap, fig.width=8, fig.height=6}
DimPlot(seurat_obj, reduction = "umap", group.by = "BRETIGEA_celltype", label = TRUE) + NoLegend()
```