---
title: "Celltype annotation with BRETIGEA"
author: "Meryem Stroosma"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    latex_engine: xelatex
    fig_caption: true
    
params:
  resolution: "0.7"
  dataset_name: "GSE138852"

# Use colours for hyperlinks
colorlinks: true
linkcolor: purple
urlcolor: purple
citecolor: purple

header-includes:
  - \usepackage{fontspec}
  - \usepackage{fancyvrb}
  - \usepackage[section]{placeins}
  - \usepackage{float} 
  - \usepackage{blindtext}
  - \setmainfont{DejaVu Sans Condensed}
  - \setsansfont{DejaVu Sans}
  - \setmonofont{DejaVu Sans Mono}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{fontsize=\footnotesize,commandchars=\\\{\}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      error = FALSE)
```
## Contents 
 - [Introduction](#introduction) 
 - [Research question](#research-question) 
 - [Packages and requirements](#packages-and-requirements)
 - [Load Seurat object](#load-seurat-object)
 - [Marker discovery per cluster](#marker-discovery-per-cluster)
 - [Marker plots](#marker-plots)
 - [Cell type annotation with BRETIGEA](#cell-type-annotation-with-bretigea)
 - [Assign labels back to Seurat](#assign-labels-back-to-seurat)
 - [UMAP visualization](#umap-visualization)
 - [Save outputs](#save-outputs)

___ 

## Packages and Requirements 
In order to run the analysis in Seurat a few packages need to be installed first. Packages and package versions are managed using ```renv```, instructions for setting up the R project with ```renv``` are given in the [README](https://github.com/ProjecticumDlerpDs/ad_transcriptomics_dge/blob/main/README.md) of the GitHub repository *ad_transcriptomics_dge*.

```{r setup_libraries, echo=FALSE, include=FALSE}
# Load the required packages
library(dplyr)
library(Seurat)
library(BRETIGEA)
library(patchwork)
library(here)
library(ggplot2)
```

```{r setup_find_seuratobject_function, echo=FALSE}
# A function was created to load a Seurat object created with any given resolution, the resolution can be given by the user upon knitting this RMarkdown. 
load_seurat_by_resolution <- function(resolution,
                                      dir = c("analyses", "seurat"),
                                      prefix = "seurat_umap_clustering_") {

  # Create a pattern to search for a filename, the clustering resolution can be given using params. The default resolution is 0.7.
  res_pattern <- paste0("res", gsub("\\.", "\\\\.", resolution), "\\.rds$")

# Search in the specified directory for Seurat objects matching the pattern.
  files <- list.files(
    path = do.call(here, as.list(dir)),
    pattern = paste0("^", prefix, res_pattern),
    full.names = TRUE
  )
# If no files are found then stop execution.
  if (length(files) == 0)
    stop("No Seurat object found for resolution ", resolution)

# If multiple files with the same pattern are found then stop execution.
  if (length(files) > 1)
    stop("Multiple Seurat objects found for resolution ", resolution)

# Print which file is used to load the Seurat object to be used for further analysis.
  message("Loading Seurat object: ", basename(files))
  readRDS(files)
}
```

## Load Seurat object
```{r Load_Seurat_object}
# Load the Seurat object with the previously made function load_seurat_by_resolution()
seurat_obj <- load_seurat_by_resolution(params$resolution)
```

## Marker discovery per cluster
Prior to analysis in BRETIGEA markers are identified per cluster in Seurat with the function ```FindAllMarkers()```.  
This step is not required for analysis in BRETIGEA however, identifying the markers in the different clusters provides relevant information about cluster composition. By verifying the expressed markers per cluster it is possible to predict which cell types are present. For annotation purposes, only positive markers were selected with the argument ```only.pos``` since cell types are mainly defined by the different genes they express. The test used to compare expression values was a Wilcoxon rank-sum test, which is a non-parametric test that doesn't assume normality. In FindAllMarkers this test compares gene expression values of a cluster to all other cells, finding genes that are uniquely expressed in that cluster. 


```{r Find_cluster_markers}
# Find markers within the clusters in seurat object with Wilcoxon rank sum test
markers_all <- FindAllMarkers(
  seurat_obj,
  only.pos = TRUE,      # Only use positive markers
  test.use = "wilcox",  # Use Wilcoxon rank sum test
  min.pct = 0.25,       # The feature must be detected in at least 25% of the cells within this cluster
  logfc.threshold = 0.25 # Minimum log fold change of 0.25 compared to other cells 
)

# Create an object with the top markers per cluster to be used in a heatmap
top_markers <- markers_all %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 10) %>%
  ungroup()

top_markers
```

## Marker plots

### Heatmap of top markers

```{r Heatmap_top_markers, fig.height=8, fig.width=8}
top5 <- markers_all %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 5) %>%
  pull(gene) %>%
  unique()

# Scale only the genes we plot (keeps it fast)
seurat_obj <- ScaleData(seurat_obj, features = top5, verbose = FALSE)

DoHeatmap(seurat_obj, features = top5, group.by = "seurat_clusters") + NoLegend()
```

### DotPlot of canonical brain markers

```{r DotPlot_canonical_markers, fig.height=6, fig.width=10}
canonical <- c(
  "RBFOX3","SNAP25","SYT1",          # neurons
  "SLC17A7",                         # excitatory
  "GAD1","GAD2",                     # inhibitory
  "AQP4","ALDH1A1","SLC1A2","SLC1A3",# astrocytes
  "MBP","PLP1","MOG",                # oligodendrocytes
  "PDGFRA","CSPG4","VCAN",           # OPC
  "P2RY12","CX3CR1","C1QA",          # microglia
  "PECAM1","VWF","KDR",              # endothelial
  "PDGFRB","RGS5","ACTA2"            # pericyte/VSMC
)

canonical_present <- canonical[canonical %in% rownames(seurat_obj)]

DotPlot(seurat_obj, features = canonical_present) + RotatedAxis()
```

### FeaturePlot (UMAP) for key markers

```{r FeaturePlot_key_markers, fig.height=7, fig.width=10}
key_markers <- c("SLC17A7","GAD1","AQP4","MBP","P2RY12","PECAM1")
key_present <- key_markers[key_markers %in% rownames(seurat_obj)]

FeaturePlot(seurat_obj, features = key_present, ncol = 3)
```

## Cell type annotation with BRETIGEA

### Create expression matrix per cluster (average expression)
```{r create_expression_matrix_per_cluster}
pb <- AverageExpression(
  seurat_obj,
  assays = "RNA",
  layer  = "data"
)$RNA

# FIX: remove the 'g' prefix added by Seurat for identities starting with numbers
colnames(pb) <- sub("^g", "", colnames(pb))
```

### Select BRETIGEA marker reference and run scoring

```{r select_Bretigea_reference}
markers <- BRETIGEA::markers_df_human_brain
genes_use <- intersect(unique(markers$marker), rownames(pb))
pb_use <- as.matrix(pb[genes_use, , drop = FALSE])
```

```{r run_bretigea_on_clusters}
ct_res <- BRETIGEA::findCells(
  inputMat = pb_use,
  markers  = markers,
  nMarker  = 20,
  method   = "PCA",
  scale    = FALSE
)
```

### Create one label per cluster (max score)
```{r Create_label_per_cluster}
scores <- as.matrix(ct_res)

# clusters are rows; cell types are columns
cluster_to_celltype <- apply(scores, 1, function(x) colnames(scores)[which.max(x)])

# Store scores + label mapping for transparency
bretigea_cluster_scores <- as.data.frame(scores)
bretigea_cluster_scores$cluster <- rownames(scores)
bretigea_cluster_scores$BRETIGEA_label <- unname(cluster_to_celltype[rownames(scores)])

bretigea_cluster_scores
```

## Assign labels back to Seurat

```{r Return_labels_to_Seuratobj}
clusters <- as.character(Idents(seurat_obj))
celltype_by_cell <- unname(cluster_to_celltype[clusters])

celltype_by_cell[is.na(celltype_by_cell)] <- "Unknown"
names(celltype_by_cell) <- colnames(seurat_obj)

# Store BRETIGEA call as a metadata column
seurat_obj <- AddMetaData(seurat_obj, celltype_by_cell, col.name = "BRETIGEA_celltype")


seurat_obj$celltype_level1 <- seurat_obj$BRETIGEA_celltype

table(seurat_obj$BRETIGEA_celltype, useNA = "ifany")
```

## UMAP visualization

```{r Plot_celltypes_umap, fig.width=8, fig.height=6}
DimPlot(seurat_obj, reduction = "umap", group.by = "BRETIGEA_celltype", label = TRUE) + NoLegend()
```