---
title: "Celltype annotation with BRETIGEA"
author: "Meryem Stroosma"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    latex_engine: xelatex
    fig_caption: true
    
params:
  resolution: "0.7"
  dataset_name: "GSE138852"

# Use colours for hyperlinks
colorlinks: true
linkcolor: purple
urlcolor: purple
citecolor: purple

header-includes:
  - \usepackage{fontspec}
  - \usepackage{fancyvrb}
  - \usepackage[section]{placeins}
  - \usepackage{float} 
  - \usepackage{blindtext}
  - \setmainfont{DejaVu Sans Condensed}
  - \setsansfont{DejaVu Sans}
  - \setmonofont{DejaVu Sans Mono}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{fontsize=\footnotesize,commandchars=\\\{\}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      error = FALSE,
                      message = FALSE)
```
## Contents 
 - [Introduction](#introduction) 
 - [Research question](#research-question) 
 - [Packages and requirements](#packages-and-requirements)
 - [Load Seurat object](#load-seurat-object)
 - [Find markers per cluster](#find-markers-per-cluster)
 - [Marker plots](#marker-plots)
 - [Cell type annotation with BRETIGEA](#cell-type-annotation-with-bretigea)
 - [Assign labels back to Seurat](#assign-labels-back-to-seurat)
 - [UMAP visualization](#umap-visualization)
 - [Save outputs](#save-outputs)

___ 
## Introduction
This RMarkdown describes analysis of the single-nucleus RNA-sequencing count matrix from [GSE138852](https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE138852&format=file&file=GSE138852%5Fcounts%2Ecsv%2Egz) using Seurat and BRETIGEA. Prior to this RMarkdown pre-processing, PCA and clustering of the count matrix should allready have been performed as described in ```02_identification_most_variable_features_in_Seurat.Rmd```.

The scope of this RMarkdown includes finding markers with Seurat, plotting markers to verify which genes are stronger expressed within clusters and celltype annotation with BRETIGEA.
BRETIGEA is short for BRain cEll Type specIfic Gene Expression Analysis, this is an R package that could be used to annotate brain cells based on the gene expression profile. Within BRETIGEA datasets for brain tissue samples are available that could be used to identify the six major brain cell types within a dataset. 

## Research question
For this project the main research question is _"How can a reliable and reproducible analysis pipeline be set up to identify cell type-specific differential gene expression in Alzheimer's disease using a filtered expression matrix generated from snRNA-seq data?"_. 

The sub-question(s) that is answered in this RMarkdown is as followed;
 - Can different cell types be distinguished from each other with BRETIGEA based on features found after sequencing RNA from the nuclei?

## Packages and Requirements 
In order to run the analysis in Seurat a few packages need to be installed first. Packages and package versions are managed using ```renv```, instructions for setting up the R project with ```renv``` are given in the [README](https://github.com/ProjecticumDlerpDs/ad_transcriptomics_dge/blob/main/README.md) of the GitHub repository *ad_transcriptomics_dge*.

```{r setup_libraries, echo=FALSE, include=FALSE}
# Load the required packages
library(dplyr)
library(Seurat)
library(BRETIGEA)
library(patchwork)
library(here)
library(ggplot2)
```

```{r setup_find_seuratobject_function, echo=FALSE}
# A function was created to load a Seurat object created with any given resolution, the resolution can be given by the user upon knitting this RMarkdown. 
load_seurat_by_resolution <- function(resolution,
                                      dir = c("analyses", "seurat"),
                                      prefix = "seurat_umap_clustering_") {

  # Create a pattern to search for a filename, the clustering resolution can be given using params. The default resolution is 0.7.
  res_pattern <- paste0("res", gsub("\\.", "\\\\.", resolution), "\\.rds$")

# Search in the specified directory for Seurat objects matching the pattern.
  files <- list.files(
    path = do.call(here, as.list(dir)),
    pattern = paste0("^", prefix, res_pattern),
    full.names = TRUE
  )
# If no files are found then stop execution.
  if (length(files) == 0)
    stop("No Seurat object found for resolution ", resolution)

# If multiple files with the same pattern are found then stop execution.
  if (length(files) > 1)
    stop("Multiple Seurat objects found for resolution ", resolution)

# Print which file is used to load the Seurat object to be used for further analysis.
  message("Loading Seurat object: ", basename(files))
  readRDS(files)
}
```

## Load Seurat object
```{r Load_Seurat_object}
# Load the Seurat object with the previously made function load_seurat_by_resolution()
seurat_obj <- load_seurat_by_resolution(params$resolution)
```

## Find markers per cluster
Prior to analysis in BRETIGEA markers are identified per cluster in Seurat with the function ```FindAllMarkers()```.  
This step is not required for analysis in BRETIGEA however, identifying the markers in the different clusters provides relevant information about cluster composition. By verifying the expressed markers per cluster it is possible to predict which cell types are present. For annotation purposes, only positive markers were selected with the argument ```only.pos``` since cell types are mainly defined by the different genes they express. The test used to compare expression values was a Wilcoxon rank-sum test, which is a non-parametric test that doesn't assume normality. In FindAllMarkers this test compares gene expression values of a cluster to all other cells, finding genes that are uniquely expressed in that cluster. 


```{r Find_cluster_markers}
# Find markers within the clusters in seurat object with Wilcoxon rank sum test
markers_all <- FindAllMarkers(
  seurat_obj,
  only.pos = TRUE,      # Only use positive markers
  test.use = "wilcox",  # Use Wilcoxon rank sum test
  min.pct = 0.25,       # The feature must be detected in at least 25% of the cells within this cluster
  logfc.threshold = 0.25, # Minimum log fold change of 0.25 compared to other cells 
  verbose = FALSE 
)

# Use the output of FindAllMarkers to create an object with the top markers per cluster to be used in a heatmap
top_markers <- markers_all %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 10) %>% # Rank the markers so the top 10 markers with highest Log fold change are used
  ungroup() # Ungroup to make sure the object behaves as it should in further analysis
```

## Marker plots

### Heatmap of top markers
In figure 1 a heatmap was plotted. This heatmap shows the top five markers per cluster. The coloured bar on top of the heatmap shows the identity of each cluster, on the y-axis of the heatmap the different markers are shown. 

```{r Heatmap_top_markers, fig.height=8, fig.width=8}
# Create a vector of the gene IDs per cluster to be used in a heatmap
top5 <- markers_all %>% 
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 5) %>% # Rank the markers so the top 5 markers with highest Log fold change are used
  pull(gene) %>%
  unique()

# Scale only the genes we plot (keeps it fast)
seurat_obj <- ScaleData(seurat_obj, features = top5, verbose = FALSE)

DoHeatmap(seurat_obj, features = top5, group.by = "seurat_clusters") + NoLegend()
```

### DotPlot of canonical brain markers

```{r DotPlot_canonical_markers, fig.height=6, fig.width=10}
dotplot_markers <- c(
  "RELN","SYT1",                     # neurons
  "SLC17A7",                         # excitatory
  "GAD1","GAD2",                     # inhibitory
  "AQP4","GFAP","ALDH1L1",           # astrocytes
  "MBP","PLP1","MOG",                # oligodendrocytes
  "PDGFRA","VCAN",                   # OPC
  "C1QA", "C1QB", "CD74", "CSF1R"   # microglia
)

dotplot_markers_present <- dotplot_markers[dotplot_markers %in% rownames(seurat_obj)]

DotPlot(seurat_obj, features = dotplot_markers_present) + RotatedAxis()
```

## Cell type annotation with BRETIGEA

### Create expression matrix per cluster (average expression)
```{r create_expression_matrix_per_cluster}
pb <- AverageExpression(
  seurat_obj,
  assays = "RNA",
  layer  = "data"
)$RNA

# FIX: remove the 'g' prefix added by Seurat for identities starting with numbers
colnames(pb) <- sub("^g", "", colnames(pb))
```

### Select BRETIGEA marker reference and run scoring

```{r select_Bretigea_reference}
markers <- BRETIGEA::markers_df_human_brain
genes_use <- intersect(unique(markers$marker), rownames(pb))
pb_use <- as.matrix(pb[genes_use, , drop = FALSE])
```

```{r run_bretigea_on_clusters}
ct_res <- BRETIGEA::findCells(
  inputMat = pb_use,
  markers  = markers,
  nMarker  = 20,
  method   = "PCA",
  scale    = FALSE
)
```

### Create one label per cluster (max score)
```{r Create_label_per_cluster}
scores <- as.matrix(ct_res)

# clusters are rows; cell types are columns
cluster_to_celltype <- apply(scores, 1, function(x) colnames(scores)[which.max(x)])

# Store scores + label mapping for transparency
bretigea_cluster_scores <- as.data.frame(scores)
bretigea_cluster_scores$cluster <- rownames(scores)
bretigea_cluster_scores$BRETIGEA_label <- unname(cluster_to_celltype[rownames(scores)])

bretigea_cluster_scores
```

## Assign labels back to Seurat

```{r Return_labels_to_Seuratobj}
clusters <- as.character(Idents(seurat_obj))
celltype_by_cell <- unname(cluster_to_celltype[clusters])

celltype_by_cell[is.na(celltype_by_cell)] <- "Unknown"
names(celltype_by_cell) <- colnames(seurat_obj)

# Store BRETIGEA call as a metadata column
seurat_obj <- AddMetaData(seurat_obj, celltype_by_cell, col.name = "BRETIGEA_celltype")


seurat_obj$celltype_level1 <- seurat_obj$BRETIGEA_celltype

table(seurat_obj$BRETIGEA_celltype, useNA = "ifany")
```

## UMAP visualization

```{r Plot_celltypes_umap, fig.width=8, fig.height=6}
DimPlot(seurat_obj, reduction = "umap", group.by = "BRETIGEA_celltype", label = TRUE) + NoLegend()
```